pirouette: the error BEAST2 makes in inferring a phylogeny

### 2019-01-29 article.tex
Processed feedback, thanks @Giappo, fix https://github.com/richelbilderbeek/pirouette/issues/58

The goal of \verb;pirouette; is to measure the inference error BEAST2
makes from a given phylogeny of which full information is provided \giovanni{is this true? do we also provide information about extinct species or we just use reconstructed tree (which always lack a lot of information)?}.

@@

\begin{lstlisting}[language=R, floatplacement=H]
install.packages("pirouette")
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H]
usethis::install_github("richelbilderbeek/pirouette")
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H]
library(pirouette)
\end{lstlisting}
Because \verb;pirouette; calls BEAST2, BEAST2 must be installed. 
This can be done from within R, using:

\begin{lstlisting}[language=R, floatplacement=H]
install_beast2()
\end{lstlisting}
For the option to select inference models,
\verb;pirouette; uses the "NS" BEAST2 package [Maturana et al.].
It can be installed from within R, using:

\begin{lstlisting}[language=R, floatplacement=H]
install_beast2_pkg("NS")
\end{lstlisting}

@@

% Need idealised birth-death tree
% From https://github.com/richelbilderbeek/pirouette_article/issues/1
% and https://github.com/richelbilderbeek/pirouette/issues/35
\begin{lstlisting}[language=R, floatplacement=H]
phylogeny <- ape::read.tree(text = "((A:4, B:4):1, (C:4, D:4):1);")
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H]
alignment_params <- create_alignment_params(
  root_sequence = create_blocked_dna(length = 1000),
  mutation_rate = 0.1
)
\end{lstlisting}

The second step is to select how an inference model is picked.
We just select the generative model, which consists in a collection of 
three elements: a site model \giovanni{do we have to explain what a site model is or what are the options?}, a clock
model of the alignment and a birth-death tree prior
to be the generative tree prior. \giovanni{would it be helpful to the reader to explicit that site and clock models are automatically taken from the generative model?}

\begin{lstlisting}[language=R, floatplacement=H]
model_select_param <- create_gen_model_select_param(
  alignment_params = alignment_params,
  tree_prior = create_bd_tree_prior()
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H]
errors <- pir_run(
  phylogeny,
  alignment_params = alignment_params,
  model_select_params = model_select_param
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H]
pir_plot(errors)
\end{lstlisting}

@@

% Need idealised birth-death tree
% From https://github.com/richelbilderbeek/pirouette_article/issues/1
% and https://github.com/richelbilderbeek/pirouette/issues/35
\begin{lstlisting}[language=R, floatplacement=H]
phylogeny <- ape::read.tree(text = "((A:4, B:4):1, (C:4, D:4):1);")
\end{lstlisting}

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. We'll re-use the alignment parameters of the previous example \giovanni{it might be nice to add an explicit ref to the alignmentparams code}.

@@

\begin{lstlisting}[language=R, floatplacement=H]
model_select_param <- create_best_model_select_param()
\end{lstlisting}

@@

We use all the same settings as the previous model:
phylogeny, alignment parameters, inference model selection,
shared inference parameters and error measuring parameters \giovanni{it might be nice to add an explicit ref to the lines of code we are referring to}.

@@

\begin{lstlisting}[language=R, floatplacement=H]
twinning_params <- create_twinning_params()
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H]
errors <- pir_run(
  phylogeny,
  twinning_params = twinning_params,
  alignment_params = alignment_params,
  model_select_params = model_select_param
)
\end{lstlisting}

>>>>

The goal of \verb;pirouette; is to measure the inference error BEAST2
makes from a given reconstructed phylogeny.

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
install.packages("pirouette")
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
usethis::install_github("richelbilderbeek/pirouette")
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
library(pirouette)
\end{lstlisting}
Because \verb;pirouette; calls BEAST2, BEAST2 must be installed. 
This can be done from within R, using:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
install_beast2()
\end{lstlisting}
For the option to select inference models,
\verb;pirouette; uses the "NS" BEAST2 package [Maturana et al.].
It can be installed from within R, using:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
install_beast2_pkg("NS")
\end{lstlisting}

@@

% Need idealised birth-death tree
% From https://github.com/richelbilderbeek/pirouette_article/issues/1
% and https://github.com/richelbilderbeek/pirouette/issues/35
\begin{lstlisting}[language=R, floatplacement=H, frame=single]
phylogeny <- ape::read.tree(text = "((A:4, B:4):1, (C:4, D:4):1);")
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
alignment_params <- create_alignment_params(
  root_sequence = create_blocked_dna(length = 1000),
  mutation_rate = 0.1
)
\end{lstlisting}

By default, an alignment is created using a Jukes-Cantor (JC) site model
and a strict clock model. A JC site model assumes that mutations
between all nucloetides are constant at an equal rate. 
A strict clock model assumes that
the mutation rate of all species is equal and constant.
Due to this, we state that the generative model for the alignment is
a JC site model and a strict clock model.

The second step is to select how an inference model is picked.
We just select the generative model, which consists in a collection of 
three elements: a site model, clock model and tree prior.
If we know how the generative model of the alignment, as we did above,
we can re-use that information. We do still need to specify 
a generative tree prior, which is the tree prior underlying the phylogeny.
In this case, we'll pick a (constant-rate) birth-death model:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
model_select_param <- create_gen_model_select_param(
  alignment_params = alignment_params,
  tree_prior = create_bd_tree_prior()
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
errors <- pir_run(
  phylogeny,
  alignment_params = alignment_params,
  model_select_params = model_select_param
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
pir_plot(errors)
\end{lstlisting}

@@

% Need idealised birth-death tree
% From https://github.com/richelbilderbeek/pirouette_article/issues/1
% and https://github.com/richelbilderbeek/pirouette/issues/35
\begin{lstlisting}[language=R, floatplacement=H, frame=single]
phylogeny <- ape::read.tree(text = "((A:4, B:4):1, (C:4, D:4):1);")
\end{lstlisting}

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. We'll re-use the alignment parameters of the previous example 
\richel{TODO: add ref to code}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
model_select_param <- create_best_model_select_param()
\end{lstlisting}

@@

We use all the same settings as the previous model:
phylogeny, alignment parameters, inference model selection,
shared inference parameters and error measuring parameters 
\richel{TODO: add ref to code}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
twinning_params <- create_twinning_params()
\end{lstlisting}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
errors <- pir_run(
  phylogeny,
  twinning_params = twinning_params,
  alignment_params = alignment_params,
  model_select_params = model_select_param
)
\end{lstlisting}
